---
title: "Comparison of same-day Sentinel-2 satellite images from different platforms"
author: "Lee, Chengfa Benjamin"
date: "`r Sys.Date()`"
output: html_document
bibliography: references.bib 
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(magrittr)
library(dplyr)
library(terra)
library(tidyterra)
library(sf)
library(ggplot2)
library(leaflet)
library(ggpubr)
library(patchwork)
```

This report explores the differences in classification output of the [ICE CREAMS model](https://github.com/BedeFfinian/ICE_CREAMS) on two images taken on the same day by Sentinel-2. 
- 22 Feb 2026: Sentinel-2A and Sentinel-2B passed over Ria de Aveiro within 9 minutes of each other.

All analyses were performed in R version 4.5.2 [@Rversion].
All work belong to the [REWRITE project](https://rewriteproject.eu/).

This investigation is a work in progress. Information may change.

# Satellite images

The overlap in coverage is noted at Ria de Aveiro. with only about 9 minutes of difference and no clouds, the light conditions are assumed to be the same.

```{r aveiro_map_preprocess}
#| echo: false
#| error: false
#| message: false
#| warning: false

shp.AV <- read_sf('data/shp/Aveiro_intertidal_GEE.shp') 
shp.AV.bb <- shp.AV %>%
  st_bbox() %>%
  st_as_sfc()

rgb.S2A <- c(rast('data/s2/S2A/T29TNF_20260222T113001_B02_10m.tif'),
             rast('data/s2/S2A/T29TNF_20260222T113001_B03_10m.tif'),
             rast('data/s2/S2A/T29TNF_20260222T113001_B04_10m.tif')) %>%
  crop(shp.AV.bb) 
rgb.S2B <- c(rast('data/s2/S2B/T29TNF_20260222T112109_B02_10m.tif'),
             rast('data/s2/S2B/T29TNF_20260222T112109_B03_10m.tif'),
             rast('data/s2/S2B/T29TNF_20260222T112109_B04_10m.tif')) %>%
  crop(shp.AV.bb)
```

```{r aveiro_map, fig.cap = "Images taken from the different Sentinel-2 platforms over Ria de Aveiro on the same day."}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 7.14

p_S2A <- ggplot() +
  geom_spatraster_rgb(data = rgb.S2A, r = 3, g = 2, b = 1, max_col_value = 5000) +
  ggtitle('Sentinel-2A')
p_S2B <- ggplot() +
  geom_spatraster_rgb(data = rgb.S2B, r = 3, g = 2, b = 1, max_col_value = 5000) +
  ggtitle('Sentinel-2B')

p_S2A + p_S2B
```

# ICE CREAMS model

Given such similar reflectance, one would assume that the ICE CREAMS classification would be highly similar. However, there are some differences.

```{r icecreams_preprocessing}
#| echo: false
#| error: false
#| message: false
#| warning: false

mat <- matrix(c(
  -0.5,  0.5,  NA,
   0.5,  1.5,  NA,
   1.5,  2.5,  NA,
   2.5,  3.5,  NA,
   3.5,  4.5,   1,
   4.5,  5.5,  NA,
   5.5,  6.5,  NA,
   6.5,  7.5,  NA,
   7.5,  8.5,  NA,
   8.5,  9.5,  NA
), ncol = 3, byrow = TRUE)

map.class.S2A <- rast('data/ice_creams/S2A_MSIL2A_20260222T113001_N0512_R037_T29TNF_20260222T155510_Pred_BFRD_V1_3.tif')
  map.class.S2A$Out_Class <- map.class.S2A$Out_Class %>%
    classify(mat, include.lowest = TRUE)
map.class.S2B <- rast('data/ice_creams/S2B_MSIL2A_20260222T112109_N0512_R037_T29TNF_20260222T165721_Pred_BFRD_V1_3.tif')
  map.class.S2B$Out_Class <- map.class.S2B$Out_Class %>%
    classify(mat, include.lowest = TRUE)

map.spc.diff <- abs(map.class.S2A$Seagrass_Cover - map.class.S2B$Seagrass_Cover) %>%
  classify(matrix(c(-0.001, 0.001, NA), ncol = 3, byrow = TRUE),
           include.lowest = TRUE)
```

```{r icecreams_map, fig.cap = "ICE CREAMS classification over the two images and the absolute difference in SPC output."}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.width: 12
#| fig.height: 8

p_S2A_icecreams <- ggplot() +
  geom_spatraster_rgb(data = rgb.S2A, r = 3, g = 2, b = 1, max_col_value = 5000) +
  geom_spatraster(data = map.class.S2A, aes(fill = Out_Class), na.rm = TRUE)  +
  scale_fill_viridis_c('Class', breaks = 1, labels = "Seagrass", na.value = "transparent") +
  ggtitle('Sentinel-2A')
p_S2B_icecreams <- ggplot() +
  geom_spatraster_rgb(data = rgb.S2B, r = 3, g = 2, b = 1, max_col_value = 5000) +
  geom_spatraster(data = map.class.S2B, aes(fill = Out_Class), na.rm = TRUE)  +
  scale_fill_viridis_c('Class', breaks = 1, labels = "Seagrass", na.value = "transparent") +
  ggtitle('Sentinel-2B')

p_spc_icecreams <- ggplot() +
  geom_spatraster_rgb(data = rgb.S2B, r = 3, g = 2, b = 1, max_col_value = 5000) +
  geom_spatraster(data = map.spc.diff, aes(fill = Seagrass_Cover), na.rm = TRUE)  +
  scale_fill_viridis_c('Abs % Diff', na.value = "transparent") +
  ggtitle('Absolute SPC difference')

p_S2A_icecreams + p_S2B_icecreams + p_spc_icecreams
```

# Comparing differences

By using the two ICE CREAMS class maps, the areas where the class maps agree and disagree with each other can be identified. Then, the spectral comparisons between the images based on these areas can be compared. As there are too many pixels in agreement, only pixels of both class maps that agree on the "seagrass" label are considered for the agreement. For the disagreement, so long as one of the class maps labels the pixel as "seagrass", these pixels were considered. 

```{r gee_preprocessing}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.cap: ICE CREAMS seagrass classification over the two images and the absolute difference in SPC output.

csv <- {
  diff <- read.csv('data/gee/Aveiro_same_day_sensor_comparison_diff_2026-02-22.csv') %>%
    select(-.geo)
    diff$agree <- 'diff'
  
  same <- csv <- read.csv('data/gee/Aveiro_same_day_sensor_comparison_same_2026-02-22.csv') %>%
    select(-.geo)
    same$agree <- 'same'
  
  rbind(diff,same)
}

wavelength <- data.frame('band' = c('B1','B2','B3','B4','B5','B6','B7','B8','B8A','B9','B11','B12'),
                         'wavelength' = c(443, 496, 560, 664, 703, 740, 782, 835, 864, 945, 1613, 2202))

csv.sensor <- csv %>%
  select(!contains('dist')) %>%
  pivot_longer(cols = !c(system.index, agree, contains('class'), contains('spc')),names_sep = "_", names_to = c('band', 'sensor')) %>%
  filter(band !='WVP') %>%
  left_join(wavelength, by = join_by(band)) 
  csv.sensor$band <- factor(csv.sensor$band, levels = c('B1','B2','B3','B4','B5','B6','B7','B8','B8A','B9','B11','B12'))
  
csv.dist <- csv %>%
  select(c(system.index, contains('dist'), agree, contains('class'), contains('spc'))) %>%
  pivot_longer(cols = !c(system.index, agree, contains('class'), contains('spc')),names_sep = "_", names_to = c('method', 'band')) %>%
  group_by(system.index, agree)

## grp is for grouping and then subsampling some of the pixels for a closer look 
## not practical to print all the results (too many pixels)
csv.sensor$grp <- csv.sensor$system.index %>%
  "/"(21) %>%
  floor()

csv.sensor <- csv.sensor %>%
  group_by(system.index, agree, band) %>%
  mutate(sensor.diff = diff(value), id = paste(system.index, sensor, agree, sep = "_")) %>% ## S2B - S2A
  ungroup()

## preprocessing for the matchup plot
csv.sensor.wide <- csv.sensor %>%
  group_by(system.index, agree, band) %>%
  select(-id) %>%
  pivot_wider(names_from = sensor, values_from = value) %>%
  ungroup()
```

## Match ups

```{r gee_class_matchup_preprocess}
#| echo: false
#| error: false
#| message: false
#| warning: false

table.class.matchup <- csv.sensor %>%
  filter(sensor == 'S2B') %>% ## remove duplicate that resulted from pivot_longer()
  select(class_S2A, class_S2B) %>%
  table() 

# add in missing classes (rows/columns)
# be very careful of matrix object type, else cbind.table might convert table into long format instead
if(sum(1:9 %in% colnames((table.class.matchup))) != 9){
  missingcol <- which(!1:9 %in% colnames((table.class.matchup)))
  replacementcol <- matrix(rep(0, nrow(table.class.matchup) * length(missingcol)), ncol = length(missingcol))
    colnames(replacementcol) <- missingcol
    row.names(replacementcol) <- row.names(table.class.matchup)
  
  table.class.matchup <- cbind(table.class.matchup, replacementcol) 
  
  table.class.matchup <- table.class.matchup[, order(colnames(table.class.matchup))]
}
if(sum(1:9 %in% row.names((table.class.matchup))) != 9){
  missingrow <- which(!1:9 %in% row.names((table.class.matchup)))
  replacementrow <- matrix(rep(0, ncol(table.class.matchup) * length(missingrow)), ncol = ncol(table.class.matchup))
    row.names(replacementrow) <- missingrow
    colnames(replacementrow) <- colnames(table.class.matchup)
  
  table.class.matchup <- rbind(table.class.matchup, replacementrow) 
  
  table.class.matchup <- table.class.matchup[order(row.names(table.class.matchup)),]
}

# rename rows/columns
labels.icecreams <- c("Bare Sand","Bare Sediment","Chlorophyta","Magnoliopsida","Microphytobenthos","Phaeophyceae","Rhodophyta","Water","Xanthophyceae")
  row.names(table.class.matchup) <- paste('S2A:', labels.icecreams)
  colnames(table.class.matchup) <- paste('S2B:', labels.icecreams)
```

`r (table.class.matchup['S2A: Magnoliopsida','S2B: Magnoliopsida']/sum(table.class.matchup['S2A: Magnoliopsida',]) * 100) %>% format(digits = 3, nsmall = 1)`% and `r (table.class.matchup['S2A: Magnoliopsida','S2B: Magnoliopsida']/sum(table.class.matchup[,'S2B: Magnoliopsida']) * 100) %>% format(digits = 3, nsmall = 1)`% of the seagrass labels were in agreement for Sentinel-2A and 2B, respectively. Of the pixel labelled as seagrass in the Sentinel-2A image, `r (table.class.matchup['S2A: Magnoliopsida','S2B: Bare Sediment']/sum(table.class.matchup['S2A: Magnoliopsida',]) * 100) %>% format(digits = 3, nsmall = 1)`% are labelled as bare sediment and `r (table.class.matchup['S2A: Magnoliopsida','S2B: Rhodophyta']/sum(table.class.matchup['S2A: Magnoliopsida',]) * 100) %>% format(digits = 3, nsmall = 1)`% as Rhodophyta in the Sentinel-2B image. As for the Sentinel-2B image, `r (table.class.matchup['S2A: Bare Sediment','S2B: Magnoliopsida']/sum(table.class.matchup[,'S2B: Magnoliopsida']) * 100) %>% format(digits = 3, nsmall = 1)`% of its seagrass labels were labelled as bare sediment and `r (table.class.matchup['S2A: Bare Sediment','S2B: Magnoliopsida']/sum(table.class.matchup[,'S2B: Magnoliopsida']) * 100) %>% format(digits = 3, nsmall = 1)`% as Rhodophyta in the Sentinel-S2A image. While it seems intuitive to suggest that Sentinel-2A is overestimating, it is also possible that Sentinel-2B might be underestimating.

```{r gee_class_matchup_table}
#| echo: false
#| error: false
#| message: false
#| warning: false
 
knitr::kable(table.class.matchup,
             format = 'pipe', 
             align = 'c',
             caption = 'ICE CREAMS classification map match up between Sentinel-2A and Sentinel-2B "seagrass" labels. Any pixels which were not labelled as "seagrass" in either image were removed from this exploration.'
             )
```

From their class probabilities, it can be seen that although the classification map of Sentinel-2A had a higher disagreement in label compared to its counterpart based on Sentinel-2B, it had a seemingly normally distributed class probability for the non-seagrass labels. For both its seagrass pixels in agreement and disagreement, the class probability has a left skew. Although using a class-probability threshold to delineate seagrass may be tempting, it can still lead to overestimation. For example, nearly 20,000 seagrass pixels from Sentinel-2A that disagree with Sentinel-2B have a class probability of 100%, compared with just over 20,000 agreeing pixels at 100% probability. If Sentinel-2B did not underestimate, relying on probability alone could therefore inflate seagrass extent. 

Correspondingly, many pixels with the Bare Sediment and Rhodophyta classifications for the Sentinel-2B image had a higher class probability, in spite of having a seagrass label from Sentinel-2A instead. There was more disagreement (or perhaps confusion?) between seagrass and bare sediments, followed by seagrass and Rhodophyta.

```{r gee_traits}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.width: 14
#| fig.height: 8
#| fig.cap: ICE CREAMS classification probability based on the classification labels of each image. Should there be a disagreement, then the label arising from the other image is the opposite (i.e. seagrass/Magnoliopsida vs non-seagrass). The red dotted line denotes the probability at 0.5. From the count, it can be seen that the Sentinel-2A image has classified more pixels as seagrass than the Sentinel-2B image.

csv.sensor$class_S2A <- csv.sensor$class_S2A %>%
  factor(levels = c(1:9), labels = labels.icecreams)
csv.sensor$class_S2B <- csv.sensor$class_S2B %>%
  factor(levels = c(1:9), labels = labels.icecreams)

p_hist_agree_classProbS2A <- csv.sensor %>%
  filter(sensor == 'S2A') %>% # remove duplicate
  ggplot(aes(x = classProb_S2A)) +
    geom_vline(xintercept = 0.5, colour = 'red', linetype = 2) +
    geom_histogram() +
    scale_x_continuous(limits = c(0,1), breaks = c(0, 0.25, 0.5, 0.75, 1)) + 
    xlab('Class Probability: Sentinel-2A') + ylab('Count') +
    facet_grid(agree ~ class_S2A, labeller = labeller(agree = c(diff = 'Disagreement on\nseagrass label', same = 'Agreement on\nseagrass label')))

p_hist_agree_classProbS2B <- csv.sensor %>%
  filter(sensor == 'S2B') %>% # remove duplicate
  ggplot(aes(x = classProb_S2B)) +
    geom_vline(xintercept = 0.5, colour = 'red', linetype = 2) +
    geom_histogram() +
    scale_x_continuous(limits = c(0,1), breaks = c(0, 0.25, 0.5, 0.75, 1)) +
    xlab('Class Probability: Sentinel-2B') + ylab('Count') +
    facet_grid(agree ~ class_S2B, labeller = labeller(agree = c(diff = 'Disagreement on\nseagrass label', same = 'Agreement on\nseagrass label')))

p_hist_agree_classProbS2A / p_hist_agree_classProbS2B
```

## Seagrass percent cover

When comparing the seagrass percent cover, based on the more positive kurtosis seen in the distribution for the Sentinel-2A classification map, it seems that more seagrass pixels with a lower seagrass cover (< ~ 35%) might have been labelled differently in the other image. This kurtosis pattern is not present for the distribution of pixels in disagreement from the Sentinel-2B classification map.

```{r gee_spc}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.cap: ICE CREAMS seagrass percent cover over seagrass-labelled pixels for both images. While both images may disagree in labels for some of the pixels, only pixels labelled as seagrass by the image of interest is represented in each graph.

p_hist_agree_spcS2A <- csv.sensor %>%
  filter(sensor == 'S2A') %>% # remove duplicate
  filter(class_S2A == "Magnoliopsida") %>%
  filter(spc_S2A > 0) %>%
  ggplot(aes(x = spc_S2A)) +
    # geom_vline(xintercept = 0.5, colour = 'red', linetype = 2) +
    geom_histogram() +
    scale_x_continuous(limits = c(0,100), breaks = c(0, 25, 50, 75, 100)) +
    xlab('Seagrass Percent Cover: Sentinel-2A') + ylab('Count') +
    facet_grid(agree ~ ., labeller = labeller(agree = c(diff = 'Disagreement on\nseagrass label', same = 'Agreement on\nseagrass label')))

p_hist_agree_spcS2B <- csv.sensor %>%
  filter(sensor == 'S2B') %>% # remove duplicate
  filter(class_S2B == "Magnoliopsida") %>%
  filter(spc_S2B > 0) %>%
  ggplot(aes(x = spc_S2B)) +
    # geom_vline(xintercept = 0.5, colour = 'red', linetype = 2) +
    geom_histogram() +
    scale_x_continuous(limits = c(0,100), breaks = c(0, 25, 50, 75, 100)) +
    xlab('Seagrass Percent Cover: Sentinel-2B') + ylab('Count') +
    facet_grid(agree ~ ., labeller = labeller(agree = c(diff = 'Disagreement on\nseagrass label', same = 'Agreement on\nseagrass label')))

p_hist_agree <- p_hist_agree_spcS2A + p_hist_agree_spcS2B + plot_layout(axes = 'collect')

# p_ranges_x <- c(ggplot_build(p_hist_agree[[1]])$layout$panel_scales_x[[1]]$range$range,
#   ggplot_build(p_hist_agree[[2]])$layout$panel_scales_x[[1]]$range$range)

p_ranges_y <- c(ggplot_build(p_hist_agree[[1]])$layout$panel_scales_y[[1]]$range$range,
                ggplot_build(p_hist_agree[[2]])$layout$panel_scales_y[[1]]$range$range)

p_hist_agree & 
  # xlim(min(p_ranges_x), max(p_ranges_x)) & 
  ylim(min(p_ranges_y), max(p_ranges_y))
```

## Spectral difference

Indeed, while the pixels that disagree show a larger range of spectral differences, by and large, their range of values and interquartile ranges are still very similar (no statistical tests performed).

Most importantly, there seems to be a substantial difference in B2 values between both images. For this difference plot, there seems to be a longer tail for both the boxplot and violin plots for pixels in disagreement versus pixels in agreement, as well as a possible red shift (no statistical tests performed).

```{r gee_boxplot, fig.cap = "Boxplot and violin plot showing the difference in band values between the two images for the pixels in agreement and disagreement of the seagrass classification."}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 7.14

p_box <- csv.sensor %>%
  filter(sensor == 'S2A') %>% # only one needed, the other is a dupe
  filter(wavelength < 1000) %>%
  ggplot(aes(x = band, y = sensor.diff/10000, colour = agree)) +
    geom_violin(aes(fill = agree)) +
    geom_boxplot() +
    scale_colour_viridis_d('Agreement', option = "H") +
    scale_fill_viridis_d('Agreement', option = "H", guide = FALSE) +
    xlab('Band') + ylab('Difference in band values')

p_box
```

The match ups suggest that the pixel values of these two platforms are highly similar, with many of the matchups having a "y = x" relationship. B1 and B2 had a slightly off-equal relationship with a gradient of 0.97 and 0.95/0.96, respectively. Interestingly, while the pixel labels in disagreement have a gradient of 1 for B3, the pixel labels in agreement have a gradient of 0.99 for B3 instead.

```{r gee_matchup, fig.cap = "Match up of the band pixels values between the two platforms for both pixels which class maps agree and disagree."}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.width: 12
#| fig.height: 8

p_matchup <- ggplot(csv.sensor.wide, aes(x = S2A, y = S2B)) +
  geom_smooth(aes(colour = agree)) + 
  geom_point(aes(colour = agree)) +
  scale_colour_viridis_d('Agreement', option = 'H') +
  stat_cor(label.x.npc = 'left', label.y.npc = 'top') +
  stat_regline_equation(label.x.npc = 'centre', label.y.npc = 'bottom', decreasing = TRUE) +
  facet_wrap(. ~ agree + band, ncol = 6)
  
p_matchup
```

Finally, a subset of pixel spectral values are inspected for their spectral shapes. While some of the pixels in disagreement have a very similar spectral shape, there is a noticeable difference in B2 values across all pixels, in agreement or otherwise. This supports the earlier box and violin plots.

```{r gee_spectra1}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.width: 12
#| fig.height: 12

p_shape1 <- csv.sensor %>%
    filter(grp == 10) %>%
    filter(wavelength < 1000) %>%
    ggplot(aes(x = wavelength, y = value, group = id, colour = sensor)) +
    geom_line(linewidth = 1) +
    facet_wrap(agree + system.index ~ ., scales = 'fixed')

p_shape1
```

```{r gee_spectra2}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.width: 12
#| fig.height: 12

p_shape2 <- csv.sensor %>%
    filter(grp == 100) %>%
    filter(wavelength < 1000) %>%
    ggplot(aes(x = wavelength, y = value, group = id, colour = sensor)) +
    geom_line(linewidth = 1) +
    facet_wrap(agree + system.index ~ ., scales = 'fixed')

p_shape2
```

```{r gee_spectra3, fig.cap = "Spectra shapes of both platforms for select pixels. 'same' and 'diff' denotes the ICE CREAMS classification pixel agreement. The number below the agreement represents the pixel of interest."}
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig.width: 12
#| fig.height: 12

p_shape3 <- csv.sensor %>%
    filter(grp == 1000) %>%
    filter(wavelength < 1000) %>%
    ggplot(aes(x = wavelength, y = value, group = id, colour = sensor)) +
    geom_line(linewidth = 1) +
    facet_wrap(agree + system.index ~ ., scales = 'fixed')

p_shape3
```

# References